Legacy Data Confounds Modern Genomics Studies

### 2018-09-04 legacy.tex
small style edits

Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP.\cite{Harris2015a}
While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
This is unexpected given the similarity of the two populations, and suggests a technical difference rather than a population structure effect. 
Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.
Low frequency variants have little power to detect Hardy-Weinberg disequilibrium, in essence we would not expect any homozygous alternates for low frequency variants.

@@

Thus this mutational signal is heavily enriched in GWAS suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques. \sgcomment{do we excluse singletons? We should quantify how much reduction in signal we see.}
%For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality. 
The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal \sgcomment{How about the others?}.
For population genetic analyses, the removal of low-mapping quality individuals appears preferable to filtering.

@@

\caption{\textbf{A} \sgcomment{Don't forget to update the axis labels!} The average quality of mapped bases for each individual per population included in the 1000 Genomes sequencing project. The earliest sequencing data is used for individuals sequenced more than once. The x-axis is ranked by the mean sequencing date per population. \textbf{B} The x-axis is sorted by the sequencing date per individual. The colours indicate the sequencing centres that produced the data for each individual and the shape indicates whether the individual belongs to Phase 1 or Phase 3 of the 1000 Genomes project. The bottom plot indicates the sequencing technologies used over time.}
\label{MapQual}
\end{figure}

\subsection{Overlap of significant SNPs}
The reverse GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3664 SNPs were significant after a meta-analysis (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.  
\sgcomment{Look at the frequency distributions of these SNPs in GnomAD}

@@

\subsection{Imputation}
The state of the art imputation servers use a combination of many databases including many that are not freely available.
From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.  
In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted genotype data from all of the 1kGP individuals to the Michigan Imputation Server, and found that all of the SNPs we identified as being associated with quality for chromosomes 1 and 2 were imputed. This suggests that the imputation reference panel likely includes individuals with low sequence quality, and the dubious variants have not been removed. 
%\sgcomment{All of them? How is it possible if we only submitted Japanese? Also, I thought that we had only submitted a couple of chromosomes.} 
%\luke{Submitted all of the 1kGP, queried the combined meta-analysis list of SNPs}
\sgcomment{It may be useful to impute some of the NAG data.}

@@

\subsection{Twins Research Human Genetics}
\sgcomment{HAd we ever discusssed that these GWAS hits were not genome-wide significant?}
This paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below significance below $ p < 10^{-5}$.
The data used was imputed genotype data using Hapmap II as the reference.
One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to data quality.\cite{}

While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified. 
Furthermore, the approach they used to impute genotype data for GWAS is very common.
%A literature search of the keywords 'GWAS Genotype Imputation' yielded 19,300 results. \sgcomment{What is this about??}

@@

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNP6_Singles.jpg}
\caption{Significant SNPs found in only one population. The x-axis is sorted by genomic position.}
\label{Singles}
\end{figure}

>>>>

Harris and Pritchard reported an excess of a 3-mer *AC${\rightarrow}$*CC mutational pattern in a portion of the Japanese individuals in the 1kGP.\cite{Harris2015a}
While trying to follow up on this observations in a larger and more recent Japanese cohort, we did not find this particular signature.
However, when comparing the allele frequencies between the Japanese individuals from the 1kGP and this larger dataset, we observed an unusually large number of single nucleotide polymorphisms (SNPs) private to one of the two groups.
This is unexpected given the similarity of the two populations, and suggests a technical difference rather than a population structure effect. 
Moreover, these mismatches were maintained after filtering for low-quality regions of the human genome and standard metrics such as Hardy-Weinberg equilibrium.

@@

Thus this mutational signal is heavily enriched in GWAS suspicious SNPs, but residual signal remains in non-significant SNPs, presumably because many rare alleles found in individuals with low-quality data remain unidentifiable using association techniques. 
%For this reason, the heterogeneously distributed signal persisted in the 1kGP Japanese even after filtering the variants that were significantly associated to quality. 
The removal of individuals with average mapping quality below 30 successfully removes the *AC${\rightarrow}$*CC signal \sgcomment{How about the others?}.
For population genetic analyses, the removal of low-mapping quality individuals appears preferable to filtering.

@@

\caption{\textbf{A} \sgcomment{Don't forget to update the axis labels!} The average mapping quality of each individual per population included in the 1000 Genomes sequencing project. The earliest sequencing data is used for individuals sequenced more than once. The x-axis is ranked by the mean sequencing date per population. \textbf{B} The x-axis is sorted by the sequencing date per individual. The colours indicate the sequencing centres that produced the data for each individual and the shape indicates whether the individual belongs to Phase 1 or Phase 3 of the 1000 Genomes project. The bottom plot indicates the sequencing technologies used over time.}
\label{MapQual}
\end{figure}

\subsection{Overlap of significant SNPs}
The reverse GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3664 SNPs were significant after a meta-analysis (taking the sum of the \sgcomment{squared?} deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.  
\sgcomment{Look at the frequency distributions of these SNPs in GnomAD}

@@

\subsection{Combined Test}
%To increase statistical power, we performed a meta-analysis, computing a combined p-value for each position by taking the sum of the \sgcomment{squared?} deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant.
%To improve on the power of our test, we performed a test by combining the deviations from the null model for each SNP for each population.
%Each population will have it's own deviation from the null for each position in the genome.
%We computed the combined p value for each position by taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant.
%This resulted in an increase in the number of SNPs reaching genome wide significance to 3664 (Figure \ref{Manhattan}).

\subsection{Imputation}
The state of the art imputation servers use a combination of many databases including many that are not freely available.
From the perspective of researchers, they act as black-box imputation machines that take observed genotypes as input and return imputed genotypes.  
In order to investigate the proportion of variants that we've identified as being associated to quality that are being imputed, we submitted genotype data from low-quality 1kGP individuals to the Michigan Imputation Server, and found that all of the SNPs we identified as being associated with quality were imputed. \sgcomment{All of them? How is it possible if we only submitted Japanese? Also, I thought that we had only submitted a couple of chromosomes.}
This suggests that the imputation reference panel likely includes individuals with low sequence quality, and the dubious variants have not been removed. \sgcomment{It may be useful to impute some of the NAG data.}

@@

\subsection{Twins Research Human Genetics}
\sgcomment{Had we ever discusssed that these GWAS hits were not genome-wide significant?}
This paper identified SNPs associated to ADHD using a GWAS on quantitative measures of ADHD symptoms.
While none of their SNPs reached significance below $ p < 10^{-8}$, they reported their results below significance below $ p < 10^{-5}$.
The data used was imputed genotype data using Hapmap II as the reference.
One of the top 25 SNPs in the GWAS on ADHD (rs6057648) was identified as being significantly associated to data quality.\cite{}

While they are clear in reporting that these SNPs are not genome wide significant, one can imagine a meta-analysis using this dataset would include the false positive SNP we identified. 
Furthermore, the approach they used to impute genotype data for GWAS is very common.
A literature search of the keywords 'GWAS Genotype Imputation' yielded 19,300 results. \sgcomment{What is this about??}

@@
